---
title: "FluxAnalysis"
author: "Larissa Verona"
date: "2026-05-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Initial Processing

### Packages 
```{r}
#usethis::use_git()
#usethis::use_github(private=T)
library(tidyverse)
library(lme4)
library(performance)
library(DHARMa)
library(viridis)
library(ggthemes)
library(ggpubr)
library(factoextra)

rm(list=ls())
```

### Flux Data
```{r}
# 1. List all csv
files <- list.files(
  path = "Data/FluxProcessed",
  pattern = "^(vvt|7q|bbt|cr|ing).+\\.csv$",
  full.names = TRUE
)

# 2. Function to deal with different formats
read_allfiles <- function(arq) {
  
  # Detectar o separador
  sep_df <- if(grepl(";", readLines(arq, n = 1))) ";" else ","
  
  # 1. Pegar apenas o nome do arquivo (sem o caminho das pastas)
  # 2. Remover a extensão .csv
  file_name <- str_remove(basename(arq), "\\.csv")
  
  # 3. Agora o split vai funcionar como esperado
  meta <- str_split(file_name, "_")[[1]]
  
  df_temp <- read.csv(arq, sep = sep_df, skip = 1) %>%
    dplyr::slice(-1) %>%
    # Agora meta[1] será "vvt" e meta[2] será "set24"
    mutate(area = meta[1], month = meta[2])
  
  # Lógica de nomes de coluna
  if ("DATE_TIME.initial_value" %in% names(df_temp)) {
    df_temp <- df_temp %>% rename(datetime = DATE_TIME.initial_value)
  }
  
  # Formatos de data
  df_temp <- df_temp %>%
    mutate(datetime = parse_date_time(datetime, orders = c("dmy HM", "ymd HMS")))
  
  return(df_temp)
}

# 3. Processing
df <- files %>% 
  map_dfr(read_allfiles) %>%
  
  mutate(across(c(3, 4, 7:14), as.numeric),
         across(c(2, 16, 17), as.factor)) %>%
  rename(point = COMMENT, 
         swc = SWC_2.mean, 
         ts = TS_2.mean, 
         fch4 = FCH4_DRY, 
         fco2 = FCO2_DRY,
         lat = LATITUDE,
         long = LONGITUDE) %>%
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  mutate(area=case_when(area =="7q"~"seteq",
         TRUE ~ as.character(area)))

str(df)
rm(files)


```

### Soil Data

```{r}
soil <- read.csv("Data/Soils/Soil.csv", sep=";") 
df <- left_join(df, soil, by=c("area", "point"))
```

### HAND Data
(Hight Above Near Drainage)

```{r}
hand <- read.csv("Data/Hand/Flux_points_HAND.csv", sep=";") %>% 
  dplyr::select("area", "point", "hand") %>% 
  group_by(area) %>%
  mutate(hand_p = (hand-min(hand))/(max(hand)-min(hand)))

df <- left_join(df, hand, by=c("area", "point")) %>% mutate(area=factor(area),
       point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6", "P7", "P8", "P9", "P10")))
```

##General Analysis

### Descriptive data

CO2
```{r}
df[which.max(df$fco2),] #max
df[which.min(df$fco2),] #min 


n <- ceiling(nrow(df) * 0.05) # number of lines representing 5%
co2_top5 <- order(df$fco2, decreasing = TRUE)[1:n]
df[co2_top5, ] %>% dplyr::select(swc, C) %>% summarise(mean(C), mean(swc))

df %>% group_by(area, point) %>% summarise(meanco2 = mean(fco2), meanch4 =  mean(fch4), sdco2 = sd(fco2), sdch4 = sd(fch4), hand_p=mean(hand_p)) %>% 
  ggplot(aes(x=hand_p, y=sdch4)) +
  geom_point() +
  geom_smooth(method="lm")+
  stat_cor()
```

CH4
```{r}
df[which.max(df$fch4),] #max
df[which.min(df$fch4),] #min 


n <- ceiling(nrow(df) * 0.05) # number of lines representing 5%
co2_top5 <- order(df$fco2, decreasing = TRUE)[1:n]
df[co2_top5, ] %>% dplyr::select(swc, C) %>% summarise(mean(C), mean(swc))

df %>% filter(fch4>0) %>% group_by(month) %>% summarise(n(), mean(C), min(C), mean(hand_p), max(hand_p))

df <- df %>% 
  mutate(
    season = case_when(
      month %in% c("set24", "set25", "jun25") ~ "dry",
      TRUE ~ "wet"
    )
  )

lm(fco2~season, data=df) %>% summary()
```


## Overall Data for Fluxes

```{r}
df_overall <- df %>%
  group_by(area, point) %>%
  summarise(meanCO2 = mean(fco2), meanCH4 =  mean(fch4),  sdCO2 = sd(fco2), sdCH4 = sd(fch4), medianCO2 = median(fco2), medianCH4=median(fch4), maxCO2 = max(fco2), maxCH4 = max(fch4), minCO2 = min(fco2), minCH4 = min(fch4), hand_p=mean(hand_p), minSWC = min(swc), maxSWC=max(swc), meanSWC= mean(swc, na.rm=T), medianSWC=median(swc, na.rm=T), sdSWC=sd(swc, na.rm=T), C=mean(C), clay=mean(clay), hand=mean(hand))
```

Means, meadians, max, min and sd
```{r}
df_overall %>% 
  ggplot(aes(x=hand_p, y=meanCO2, color=C)) +
  geom_point()+
  #geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=medianCO2, color=C)) +
  geom_point()+
  #geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)



df_overall %>% 
  ggplot(aes(x=hand_p, y=sdCO2, color=meanSWC, size=C)) +
  geom_point()+
  #geom_smooth(method="lm")+
  #stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=maxCO2, color=C)) +
  geom_point()+
#  geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=minCO2, color=C)) +
  geom_point()+
#  geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)


df_overall %>% 
  ggplot(aes(x=hand_p, y=meanCH4, color=C)) +
  geom_point()+
#  geom_smooth()+
  stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=medianCH4, color=C)) +
  geom_point()+
#  geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=sdCH4, color=meanSWC, size=C)) +
  geom_point()+
#  geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=minCH4, color=C)) +
  geom_point()+
#  geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)

df_overall %>% 
  ggplot(aes(x=hand_p, y=maxCH4, color=C)) +
  geom_point()+
#  geom_smooth()+
  #stat_cor()+
  facet_wrap(~area)
```

###Predicted annual emission

```{r}
library(RColorBrewer)

df %>% 
  filter(fco2 < 12) %>% 
  ggplot(aes(x = month, y = fco2,
             color = factor(point),
             group = point)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_color_brewer(palette = "RdYlBu") +
  facet_wrap(~area)

df %>% 
  filter(fco2 < 12) %>% 
  ggplot(aes(x = month, y = fch4,
             color = factor(point),
             group = point)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_color_brewer(palette = "RdYlBu") +
  facet_wrap(~area)
```

Calculating the area below the curves

##### For CO2

1. Sampling date as numeric variable (counting days)
```{r}
date <- read.csv("Data/FluxProcessed/data_coleta_pts.csv", sep=";") %>%
  mutate(data = as.Date(data, format = "%d/%m/%Y"))

df_annual <- df %>%
  left_join(
    date %>% 
      dplyr::select(area, point, data, month),
    by = c("area", "point", "month")
  ) %>%
  group_by(area, point) %>%
  mutate(
    day_num = as.numeric(data - min(data, na.rm = TRUE))
  ) %>%
  ungroup()

```

2. Loess model
```{r}
library(purrr)
library(pracma)

# Ajuste dos modelos
loess_models <- df_annual %>%
  group_by(area, point) %>%
  summarise(
    model = list(
      loess(fco2 ~ day_num,
            data = cur_data())
    ),
    .groups = "drop"
  )

# Predições suaves
pred_data <- loess_models %>%
  left_join(
    df_annual %>%
      group_by(area, point) %>%
      summarise(
        min_day = min(day_num),
        max_day = max(day_num),
        .groups = "drop"
      ),
    by = c("area", "point")
  ) %>%
  mutate(
    newdata = map2(
      min_day,
      max_day,
      ~ tibble(
        day_num = seq(.x, .y, length.out = 200)
      )
    ),
    
    pred = map2(model, newdata,
                ~ predict(.x, newdata = .y))
  )
# Transformar em dataframe longo
pred_long <- pred_data %>%
  mutate(
    df_pred = map2(newdata, pred,
                   ~ tibble(
                     day_num = .x$day_num,
                     pred = .y
                   ))
  ) %>%
  dplyr::select(area, point, df_pred) %>%
  unnest(df_pred)
```

3. Verifying
```{r}
ggplot() +

  geom_point(
    data = df_annual,
    aes(x = day_num,
        y = fco2
        #color = point)
  )) +

  geom_line(
    data = pred_long,
    aes(x = day_num,
        y = pred,
        #color = point,
        group = interaction(area, point)),
    linewidth = 1
  ) +

  facet_wrap(~ interaction(area, point)) +

  theme_minimal()
```

4. Area above curve
```{r}

areas_co2 <- pred_long %>%
  group_by(area, point) %>%
  summarise(
    emission_umol = trapz(day_num, pred) * 86400, #seconds_per_day,
    .groups = "drop"
  )

```


##### For CH4


1. Loess model
```{r}

# Ajuste dos modelos
loess_models_ch4 <- df_annual %>%
  group_by(area, point) %>%
  summarise(
    model = list(
      loess(fch4 ~ day_num,
            data = cur_data())
    ),
    .groups = "drop"
  )

# Predições suaves
pred_data_ch4 <- loess_models_ch4 %>%
  left_join(
    df_annual %>%
      group_by(area, point) %>%
      summarise(
        min_day = min(day_num),
        max_day = max(day_num),
        .groups = "drop"
      ),
    by = c("area", "point")
  ) %>%
  mutate(
    newdata = map2(
      min_day,
      max_day,
      ~ tibble(
        day_num = seq(.x, .y, length.out = 200)
      )
    ),
    
    pred = map2(model, newdata,
                ~ predict(.x, newdata = .y))
  )
# Transformar em dataframe longo
pred_long_ch4 <- pred_data_ch4 %>%
  mutate(
    df_pred = map2(newdata, pred,
                   ~ tibble(
                     day_num = .x$day_num,
                     pred = .y
                   ))
  ) %>%
  dplyr::select(area, point, df_pred) %>%
  unnest(df_pred)
```

3. Verifying
```{r}
ggplot() +

  geom_point(
    data = df_annual,
    aes(x = day_num,
        y = fch4
        #color = point)
  )) +

  geom_line(
    data = pred_long_ch4,
    aes(x = day_num,
        y = pred,
        #color = point,
        group = interaction(area, point)),
    linewidth = 1
  ) +

  facet_wrap(~ interaction(area, point)) +

  theme_minimal()
```

4. Area above curve
```{r}

areas_ch4 <- pred_long_ch4 %>%
  group_by(area, point) %>%
  summarise(
    emission_nmol = trapz(day_num, pred) * 86400, #seconds_per_day,
    .groups = "drop"
  )

```

```{r}
areas_co2 <- areas_co2 %>% rename(emission_umolCO2 = emission_umol)
areas_ch4 <- areas_ch4 %>% rename(emission_nmolCH4 = emission_nmol)

df_overall <- df_overall %>% 
  left_join(areas_co2, by=c("area", "point")) %>%
  left_join(areas_ch4, by=c("area", "point")) %>%
  mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6", "P7", "P8", "P9", "P10")))

df_overall %>%
  ggplot(aes(x=hand_p, y=emission_umolCO2, color=meanSWC, size=C))+
  geom_point()+
  geom_hline(yintercept = 0)+
  facet_wrap(~area) 

df_overall %>%
  ggplot(aes(x=point, y=emission_nmolCH4, color=meanSWC, size=C))+
  geom_point()+
  #geom_smooth(method="lm")+
  geom_hline(yintercept = 0)+
  facet_wrap(~area) 

```

## Relation Carbon and Water

```{r}
df_overall %>% ggplot(aes(x=medianSWC, y=C)) +
  geom_point() +
  stat_cor() +
  geom_smooth(method="lm")

df_overall %>% ggplot(aes(x=meanSWC, y=sqrt(C))) +
  geom_point(aes(color=point))+
  stat_cor() +
  geom_smooth(method="lm") 

df_overall %>% ggplot(aes(x=sdSWC, y=C)) +
  geom_point()+
  stat_cor() +
  geom_smooth(method="lm")
```

Getting a PCA

```{r}
df_overall <- df_overall %>%
  mutate(
    sqrtC = sqrt(C),
    label = interaction(area, point, sep = "_")
  )

# PCA
pca <- prcomp(
  df_overall[, c("sqrtC", "meanSWC", "hand_p")],
  scale. = TRUE
)

# Scree plot
fviz_eig(pca, addlabels = TRUE)

# Biplot com labels personalizados
fviz_pca_biplot(
  pca,
  label = "var",                 # mostra nomes das variáveis
  habillage = df_overall$area,   # opcional: colore por área
  addEllipses = FALSE,
  repel = TRUE
) +
  geom_text(
    aes(
      x = pca$x[,1],
      y = pca$x[,2],
      label = df_overall$label
    ),
    size = 3
  )

df_overall$PC1 <- pca$x[,1]
df_overall$PC2 <- pca$x[,2]
```


```{r}
df_overall %>% 
  ggplot(aes(y=PC1, x=sqrtC)) +
  geom_point()+
  stat_cor()+
  geom_smooth(method="lm")

df_overall %>% 
  ggplot(aes(y=PC1, x=meanSWC)) +
  geom_point()+
  stat_cor()+
  geom_smooth(method="lm")

df_overall %>% 
  ggplot(aes(y=PC1, x=hand)) +
  geom_point()+
  stat_cor()+
  geom_smooth(method="lm")

df_overall %>% 
  ggplot(aes(y=PC1, x=hand_p)) +
  geom_point()+
  stat_cor()+
  geom_smooth(method="lm")


```

Clustering by this axis

```{r}
min_pc1 <- min(df_overall$PC1)
max_pc1 <- max(df_overall$PC1)

breaks <- seq(
  from = min_pc1,
  to = max_pc1,
  length.out = 4
)

breaks

df_class <- df_overall %>%
  ungroup() %>%
  mutate(
    PC1_group = cut(
      PC1,
      breaks = breaks,
      labels = 1:3,
      include.lowest = TRUE
    )
  )

df <- left_join(df_class, df, by = c("area", "point"))

```

```{r}
df_class2 <- df_overall %>%
  ungroup() %>%
  mutate(PC1_decil = ntile(PC1, 10)) %>%
  mutate(
    PC1_class2 = case_when(
      PC1_decil %in% c(1, 2, 3, 4) ~ "Low",
      PC1_decil %in% c(9, 10) ~ "High",
      TRUE ~ "Intermediate"
    )
  ) %>%
  mutate(PC1_decil = factor(PC1_decil),
         PC1_class2 = factor(PC1_class2, 
                             levels = c("Low", "Intermediate", "High"))) %>%
  dplyr::select(area, point, PC1_decil, PC1_class2, PC1, PC2, label)

df <- left_join(df_class2, df, by = c("area", "point"))


library(ggrepel)

df_class %>% 
  ggplot(aes(y=label, x=PC1, color=PC1_group)) +
  geom_point()

df_class %>% ggplot(aes(x=PC1_group, y=C)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=meanSWC)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=minSWC)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=maxSWC)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=hand_p)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=hand)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=meanCO2)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=meanCH4)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=minCO2)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=minCH4)) +
  geom_boxplot()
  df_class %>% ggplot(aes(x=PC1_group, y=maxCO2)) +
  geom_boxplot()
df_class %>% ggplot(aes(x=PC1_group, y=maxCH4)) +
  geom_boxplot()

df_class %>% mutate(point2 = interaction(area,point)) %>% 
  ggplot(aes(x=PC1, y=point, color=PC1_group))+
  geom_point()+
   geom_text(aes(label = point),
            hjust = -0.2)+
  facet_wrap(~area)
  

```

Ploting

```{r}
df_cluster_means <- df %>%
  group_by(PC1_group) %>%
  summarise(CH4_mean = mean(fch4), na.rm = TRUE,
            CO2_mean = mean(fco2), na.rm = TRUE)

df %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>% 
  mutate(point2 = paste(area, point)) %>%
    ggplot(aes(x = month, y = fch4, color = swc,  group = point2)) + 
  theme_clean()+
  geom_point (alpha=0.8) +
  geom_line(alpha = 0.5) + # Adiciona as linhas conectando os pontos +
  geom_smooth(aes(group = 1), method = "loess", color = "red", linetype = "dashed", se = T) +
  geom_hline(data = df_cluster_means, 
             aes(yintercept = CH4_mean), 
             color = "black", 
             linewidth = 0.5) +
  geom_hline(aes(yintercept = 0),
             color = "black", 
             linewidth = 0.5,
             linetype = "dashed"
          ) +
  facet_wrap(~PC1_group) +
     ggtitle("Clusters Annual Emissions") +
  ylab(expression(CH[4]~(nmol ~ m^{-2} * s^{-1})))+
  xlab("") +
  labs(color="SWC (%)")


df %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>% 
  mutate(point2 = paste(area, point)) %>%
  ggplot(aes(x = month, y = fco2, color = swc,  group = point2)) + 
  theme_clean()+
  geom_point ( alpha=0.8) +
  geom_line(alpha = 0.5) + # Adiciona as linhas conectando os pontos +
  geom_smooth(aes(group = 1), method = "loess", color = "red", linetype = "dashed", se = T) +
  geom_hline(data = df_cluster_means, 
             aes(yintercept = CO2_mean), 
             color = "black", 
             linewidth = 0.5) +
  geom_hline(aes(yintercept = 0),
             color = "black", 
             linewidth = 0.5,
             linetype = "dashed"
          ) +
  facet_wrap(~PC1_group) +
     ggtitle("Clusters Annual Emissions") +
  ylab(expression(CO[2]~(umol ~ m^{-2} * s^{-1})))+
  xlab("") +
  labs(color="SWC (%)")
```


```{r}
df %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  ggplot(aes(x=month, y=fco2)) +
  geom_boxplot()+
  stat_compare_means()+
  facet_wrap(~PC1_group)

df %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  ggplot(aes(x=month, y=fch4)) +
  geom_boxplot()+
  #geom_jitter()+
  stat_compare_means()+
  geom_hline(yintercept = 0)+
  facet_wrap(~PC1_group)
```


Testing

```{r}
#CO2
library(emmeans)
library(multcomp)

mod1 <- aov(log(fco2+1) ~ month*PC1_group, data=df)
performance::check_model(mod1)
summary(mod1)

emm1 <- emmeans(mod1, ~ month | PC1_group)
pairs(emm1, adjust = "tukey") %>% as.data.frame() %>% filter(p.value<0.05)

letters1 <- cld(emm1, adjust = "tukey", Letters = letters) %>%
  as.data.frame() %>%
  mutate(.group = gsub(" ", "", .group))

posicoes1 <- df %>%
  group_by(month, PC1_group) %>%
  summarise(
    ymax = max(log(fco2 + 1), na.rm = TRUE),
    .groups = "drop"
  )

letters1 <- left_join(letters1, posicoes1,
                         by = c("month", "PC1_group"))

ggplot(df,
       aes(x = month,
           y = log(fco2 + 1))) +

  geom_boxplot(position = position_dodge(width = 0.8)) +

  geom_text(data = letters1,
            aes(y = ymax + 0.2,
                label = .group,
                group = PC1_group),
            position = position_dodge(width = 0.8),
            size = 5) +

  theme_bw() +
  facet_wrap(~PC1_group)
```

CH4
```{r}
mod2 <- aov(log(fch4+5) ~ month*PC1_group, data=df)
performance::check_model(mod2)
summary(mod2)

emm2 <- emmeans(mod2, ~ month | PC1_group)
pairs(emm2, adjust = "tukey") %>% as.data.frame() %>% filter(p.value<0.05)

letters2 <- cld(emm2, adjust = "tukey", Letters = letters) %>%
  as.data.frame() %>%
  mutate(.group = gsub(" ", "", .group))

posicoes2 <- df %>%
  group_by(month, PC1_group) %>%
  summarise(
    ymax = max(log(fch4 + 5), na.rm = TRUE),
    .groups = "drop"
  )

letters2 <- left_join(letters2, posicoes2,
                         by = c("month", "PC1_group"))

ggplot(df,
       aes(x = month,
           y = log(fch4+5))) +

  geom_boxplot(position = position_dodge(width = 0.8)) +

  geom_text(data = letters2,
            aes(y = ymax + 0.2,
                label = .group,
                group = PC1_group),
            position = position_dodge(width = 0.8),
            size = 5) +

  theme_bw() +
  facet_wrap(~PC1_group)
```

##Plots

Soil Moisture
```{r}

a <- 
  df %>% 
  ggplot(aes(x=swc, y=fco2)) +
  geom_point(aes(color=C)) + 
  geom_smooth() +
  theme_bw() +
  ylab(expression("CO"[2]~" Flux (µmol/m²s)")) +
  xlab("Soil Moisture (m³/m³)") +
  labs(color="Carbon Content (%)")+
  theme(axis.text = element_text(size=10),
        axis.title = element_text(size=12)) +
  scale_color_viridis(option="inferno",direction = -1)


b <- 
  df %>% 
  #filter (C < 10) %>%
  ggplot(aes(x=swc, y=fch4)) +
  geom_point(aes(color=C)) + 
  geom_smooth() +
  theme_bw() +
  geom_hline(yintercept=0, linetype="dotted")+
  ylab(expression("CH"[4]~" Flux (nmol/m²s)")) +
  xlab("Soil Moisture (m³/m³)") +
  labs(color="Carbon Content (%)")+
  theme(axis.text = element_text(size=10),
        axis.title = element_text(size=12)) +
  scale_color_viridis(option="inferno",direction = -1) #+
  #ggtitle("Carbon < 10%")

c <- ggpubr::ggarrange(a,b, common.legend = F)
```

