---
title: "AnalisesModelosSEM"
author: "Larissa Verona"
date: "2026-03-27"
output: html_document
---
##Initial Processing

### Packages 
```{r}
#usethis::use_git()
#usethis::use_github(private=T)
library(piecewiseSEM)
library(tidyverse)
library(lme4)
library(performance)
library(DHARMa)
rm(list=ls())
```

### Flux Data
```{r}
# 1. List all csv
files <- list.files(
  path = "Data/FluxProcessed",
  pattern = "^(vvt|7q|bbt|cr|ing).+\\.csv$",
  full.names = TRUE
)

# 2. Function to deal with different formats
read_allfiles <- function(arq) {
  
  # Detectar o separador
  sep_df <- if(grepl(";", readLines(arq, n = 1))) ";" else ","
  
  # 1. Pegar apenas o nome do arquivo (sem o caminho das pastas)
  # 2. Remover a extensão .csv
  file_name <- str_remove(basename(arq), "\\.csv")
  
  # 3. Agora o split vai funcionar como esperado
  meta <- str_split(file_name, "_")[[1]]
  
  df_temp <- read.csv(arq, sep = sep_df, skip = 1) %>%
    slice(-1) %>%
    # Agora meta[1] será "vvt" e meta[2] será "set24"
    mutate(area = meta[1], month = meta[2])
  
  # Lógica de nomes de coluna
  if ("DATE_TIME.initial_value" %in% names(df_temp)) {
    df_temp <- df_temp %>% rename(datetime = DATE_TIME.initial_value)
  }
  
  # Formatos de data
  df_temp <- df_temp %>%
    mutate(datetime = parse_date_time(datetime, orders = c("dmy HM", "ymd HMS")))
  
  return(df_temp)
}

# 3. Processing
df <- files %>% 
  map_dfr(read_allfiles) %>%
  
  mutate(across(c(3, 4, 7:14), as.numeric),
         across(c(2, 16, 17), as.factor)) %>%
  rename(point = COMMENT, 
         swc = SWC_2.mean, 
         ts = TS_2.mean, 
         fch4 = FCH4_DRY, 
         fco2 = FCO2_DRY,
         lat = LATITUDE,
         long = LONGITUDE) %>%
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  mutate(area=case_when(area =="7q"~"seteq",
         TRUE ~ as.character(area)))

str(df)
rm(files)
```

### Soil Data

```{r}
soil <- read.csv("Data/Soils/Soil.csv", sep=";") 
df <- left_join(df, soil, by=c("area", "point"))
```

### HAND Data
(Hight Above Near Drainage)

```{r}
hand <- read.csv("Data/Hand/Flux_points_HAND.csv", sep=";") %>% 
  dplyr::select("area", "point", "hand") %>% 
  group_by(area) %>%
  mutate(hand_p = (hand-min(hand))/(max(hand)-min(hand)))

df <- left_join(df, hand, by=c("area", "point")) %>% mutate(area=factor(area),
       point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6", "P7", "P8", "P9", "P10")))
```

### Preciptation Data

```{r}
prep <- read.csv("Data/Preciptation/Prep_CHIRPS.csv", sep=";") %>%
  mutate(
    month = str_replace(month, "/", ""),
    point = ponto,
    area = ifelse(area == "7q", "seteq", area)
  ) %>%
  rename(prep_wk = acum_semana,
         prep_1mt = acum_1_mes,
         prep_2mt = acum_2_meses,
         prep_3mt = acum_3_meses,
         prep_4mt = acum_4_meses,
         prep_day = precip_dia) %>%
  dplyr::select(-c(data))

df <- left_join(df, prep, by=c("area", "point", "month"))
```

## Preliminar Models

Testing the preliminar models individually to see patterns

Correlations
```{r}
df_cor <- df %>%
  dplyr::select(swc, ts, fch4, fco2, N, C, sand, silt, clay, hand, hand_p, prep_wk, prep_1mt, prep_2mt, prep_3mt, prep_4mt, prep_wk, prep_day) %>%  drop_na() 

corr <- round(cor(df_cor), 2)

ggcorrplot::ggcorrplot(corr, 
           hc.order = TRUE,           
           type = "lower",           
           lab = TRUE,                
           lab_size = 3,              
           method = "circle",         #
           colors = c("#E46726", "white", "#6D9EC1"), 
           title = "Matriz de Correlação: Fluxos e Variáveis Ambientais",
           ggtheme = theme_minimal())

#Observações interessantes aqui: Carbono relacionado a hand e agua; swc relacionado a hand e preciptações, ch4 influenciado por ts, precip, e swc; co2 influenciado por hand, clay, N e C (efeito não é unico em todos os pontos)
```

Verifying the normality
```{r}
hist(df$swc)
hist(df$C)
hist(log(df$C))
shapiro.test(log(df$C)) #não ficou normal pelo teste
hist(df$prep_2mt)
shapiro.test(df$prep_2mt) 
hist(df$hand_p)
hist(df$ts)
shapiro.test(df$ts)
hist(df$fco2)
hist(log(df$fco2+1))
shapiro.test(log(df$fco2+1)) #aqui sim, vale a pena transformar
hist(df$fch4)
shapiro.test(df$fch4)
hist(log(df$fch4+5))
shapiro.test(log(df$fch4+5)) #nao muda nada para ch4

```


Models

For SWC
```{r}

mod_swc1 <- lmer(swc ~ scale(prep_2mt) * scale(hand) + (1|area/point), data=df)
check_model(mod_swc1)
simulateResiduals(fittedModel = mod_swc1) %>% plot
sjPlot::tab_model(mod_swc1)
sjPlot::plot_model(mod_swc1, type="pred", show.data = T)

#max-min scaled hand
mod_swc2 <- lmer(swc ~ scale(prep_2mt) * scale(hand_p) + (1|area/point), data=df)
check_model(mod_swc2)
simulateResiduals(fittedModel = mod_swc2) %>% plot
sjPlot::tab_model(mod_swc2)
sjPlot::plot_model(mod_swc2, type="pred", show.data = T)

anova(mod_swc1, mod_swc2) #modelo com pred_p é melhor. 

#Only prep
mod_swc3 <- lmer(swc ~ scale(prep_2mt) + (1|area/point), data=df)
check_model(mod_swc3)
simulateResiduals(fittedModel = mod_swc3) %>% plot
sjPlot::tab_model(mod_swc3)
sjPlot::plot_model(mod_swc3, type="pred", show.data = T)

#Only HAND
mod_swc4 <- lmer(swc ~ scale(hand_p) + (1|area/point), data=df)
check_model(mod_swc4)
simulateResiduals(fittedModel = mod_swc4) %>% plot
sjPlot::tab_model(mod_swc4)
sjPlot::plot_model(mod_swc4, type="pred", show.data = T)

#Null model
mod_swc5 <- lmer(swc ~ (1|area/point), data=df)
check_model(mod_swc5)
simulateResiduals(fittedModel = mod_swc5) %>% plot
sjPlot::tab_model(mod_swc5)
```

There is a strong relationship between swc and precipitation, and a weaker relationship with hand_p, but the models do not follow the assumptions.

For Carbon
```{r}
#complete model
mod_c1 <- lmer(log(C) ~ scale(hand_p)*scale(swc)+ (1|area), data=df)
check_model(mod_c1)
simulateResiduals(fittedModel = mod_c1) %>% plot()
sjPlot::tab_model(mod_c1)
sjPlot::plot_model(mod_c1, type="pred", show.data = T, terms="hand_p [all]")

#only hand_p
mod_c2 <- lmer(log(C) ~ scale(hand_p)+ (1|area), data=df)
check_model(mod_c2)
simulateResiduals(fittedModel = mod_c2) %>% plot()
sjPlot::tab_model(mod_c2)
sjPlot::plot_model(mod_c2, type="pred", show.data = T)

#only swc
mod_c3 <- lmer(log(C) ~ scale(swc)+ (1|area), data=df)
check_model(mod_c3)
simulateResiduals(fittedModel = mod_c3) %>% plot()
sjPlot::tab_model(mod_c3)
sjPlot::plot_model(mod_c3, type="pred", show.data = T)

#null model
mod_c4 <- lmer(log(C) ~ (1|area), data=df)
check_model(mod_c4)
simulateResiduals(fittedModel = mod_c4) %>% plot()
sjPlot::tab_model(mod_c4) #low r² alone
```

For temperature
```{r}
mod_temp1 <- lmer(ts ~ scale(swc)*scale(prep_2mt) + (1|point2), data = df %>% mutate(point2=paste(area, point)))
check_model(mod_temp1)
simulateResiduals(fittedModel = mod_temp1) %>% plot()
sjPlot::tab_model(mod_temp1)
sjPlot::plot_model(mod_temp1, type="pred", show.data = T)


mod_temp2 <- lmer(ts ~ (1|area), data = df)
check_model(mod_temp2)
simulateResiduals(fittedModel = mod_temp2) %>% plot()
sjPlot::tab_model(mod_temp2)
#estrutra de grupos explica bem pouco sozinha
```

For texture
```{r}
mod_text1 <- lmer(log(clay) ~ hand_p + (1|area), data=df)
check_model(mod_text1)
simulateResiduals(fittedModel = mod_text1) %>% plot()
sjPlot::tab_model(mod_text1)
# ual, que r² condicional. Esperava menos, é tudo grupo.

mod_text2 <- lmer(clay ~ (1|area), data=df)
check_model(mod_text2)
simulateResiduals(fittedModel = mod_text2) %>% plot()
sjPlot::tab_model(mod_text2)
```

For CO2 fluxes
```{r}
mod_co21 <- lmer(log(fco2+1) ~ scale(swc)*scale(C)*scale(ts)*scale(clay)+(swc|point2), data=df%>% mutate(point2 = paste(area, point)))
check_model(mod_co21)
simulateResiduals(fittedModel = mod_co21) %>% plot()
sjPlot::tab_model(mod_co21)


#That's great, super responsive model It needs simplification, but that's okay. I'll make individual templates.

mod_co22 <- lmer(log(fco2+1) ~ scale(swc)+(swc|point2), data=df%>% mutate(point2 = paste(area, point)))
check_model(mod_co22)
simulateResiduals(fittedModel = mod_co22) %>% plot()
sjPlot::tab_model(mod_co22) #swc explica bem pouco

mod_co23 <- lmer(log(fco2+1) ~ scale(C)+(1|area/point), data=df)
check_model(mod_co23)
simulateResiduals(fittedModel = mod_co23) %>% plot()
sjPlot::tab_model(mod_co23) #C sozinho explica bem pouco

mod_co24 <- lmer(log(fco2+1) ~ scale(ts)+(1|area/point), data=df)
check_model(mod_co24)
simulateResiduals(fittedModel = mod_co24) %>% plot()
sjPlot::tab_model(mod_co24) #também explica pouco mas ta super ajustado

mod_co25 <- lmer(log(fco2+1) ~ scale(clay)+(1|area), data=df)
check_model(mod_co25)
simulateResiduals(fittedModel = mod_co25) %>% plot()
sjPlot::tab_model(mod_co25)

```

For CH4 flux
```{r}
mod_ch41 <- lmer(log(fch4+5) ~ scale(swc)*scale(C)*scale(ts)*scale(clay)+(1|point2), data=df%>% mutate(point2 = paste(area, point)))
check_model(mod_ch41)
simulateResiduals(fittedModel = mod_ch41) %>% plot()
sjPlot::tab_model(mod_ch41)

#It explains little and is quite poorly adjusted. Let's look at them individually.

mod_ch42 <- lmer(log(fch4+5) ~ scale(swc)+(1|point2), data=df%>% mutate(point2 = paste(area, point)))
check_model(mod_ch42)
simulateResiduals(fittedModel = mod_ch42) %>% plot()
sjPlot::tab_model(mod_ch42) #swc explica bem, mas o ajuste ta ruim

mod_ch43 <- lmer(log(fch4+5) ~ scale(C)+(1|area/point), data=df)
check_model(mod_ch43)
simulateResiduals(fittedModel = mod_ch43) %>% plot()
sjPlot::tab_model(mod_ch43) #C sozinho explica bem pouco e quase não é significativo

mod_ch44 <- lmer(log(fch4+5) ~ scale(ts)+(1|area/point), data=df)
check_model(mod_ch44)
simulateResiduals(fittedModel = mod_ch44) %>% plot()
sjPlot::tab_model(mod_ch44) #também explica pouco 

mod_ch45 <- lmer(log(fch4+5) ~ scale(clay)+(1|area), data=df)
check_model(mod_ch45)
simulateResiduals(fittedModel = mod_ch45) %>% plot()
sjPlot::tab_model(mod_ch45) #nao explica

```


## PieceWise SEM model

For CO2 fluxes
```{r}

df_psem <- df %>% mutate(point2 = paste(area, point))  %>% dplyr::select(area, point,point2, prep_2mt, hand_p, swc, C, ts, sand, clay, fco2, fch4, month) %>% mutate(fco2 = log(fco2+1), fch4 = log(fch4+5), C=log(C)) %>% group_by(point2) %>%
  mutate(swc_mean = mean(swc)) %>% mutate(swc_resid = swc-swc_mean) %>%
  ungroup() %>%  na.omit() 

psem_co2 <- psem(
  mod1 <- lmer(swc ~ prep_2mt+hand_p+clay+ (1|area), data=df_psem),
  mod2 <- lmer(C ~  swc_mean+hand_p+clay+ (1|area), data=df_psem),
  mod3 <- lmer(ts ~  swc+prep_2mt + (1|area), data=df_psem),
 # mod4 <- lmer(clay ~ hand_p + swc_mean + (1|area), data=df_psem), #I excluded clay because it is completely explained by the random element. Therefore, it is an exogenous variable.
  mod5 <- lmer(fco2 ~ swc*ts+C+clay + (swc|area/point), data=df_psem),
  swc_mean %~~% swc,
  data=df_psem)


summary(psem_co2)

plot(psem_co2)
```

For CH4 fluxes
```{r}
psem_ch4 <- psem(
  mod1 <- lmer(swc ~ prep_2mt+hand_p+clay+ (1|area), data=df_psem),
  mod2 <- lmer(C ~  swc_mean+hand_p+clay+ (1|area), data=df_psem),
  mod3 <- lmer(ts ~  swc+prep_2mt + (1|area), data=df_psem),
 # mod4 <- lmer(clay ~ hand_p + swc_mean + (1|area), data=df_psem), #I excluded clay because it is completely explained by the random element. Therefore, it is an exogenous variable.
  mod5 <- lmer(fch4 ~ prep_2mt+swc*ts+C+clay + (1|area/point), data=df_psem),
  swc_mean %~~% swc,
  data=df_psem)


summary(psem_ch4)

plot(psem_ch4)
```



For CH4 fluxes
```{r}
psem_ch4 <- psem(
  mod1 <- lmer(swc ~ prep_2mt+hand_p+clay+ (1|area), data=df_psem),
  mod2 <- lmer(C ~  swc_mean+hand_p+clay+ (1|area), data=df_psem),
  mod3 <- lmer(ts ~  swc+prep_2mt + (1|area), data=df_psem),
 # mod4 <- lmer(clay ~ hand_p + swc_mean + (1|area), data=df_psem), #I excluded clay because it is completely explained by the random element. Therefore, it is an exogenous variable.
  mod5 <- lmer(fch4 ~ prep_2mt+swc*ts+C+clay + (1|area/point), data=df_psem),
  swc_mean %~~% swc,
  data=df_psem)


summary(psem_ch4)

plot(psem_ch4)
```

### Without peat points

```{r}
df_psem2 <- df %>% filter(C<20) %>% mutate(point2 = paste(area, point))  %>% dplyr::select(area, point,point2, prep_2mt, hand_p, swc, C, ts, sand, clay, fco2, fch4, month) %>% mutate(fco2 = log(fco2+1), fch4 = log(fch4+5), C=log(C)) %>% group_by(point2) %>%
  mutate(swc_mean = mean(swc)) %>% mutate(swc_resid = swc-swc_mean) %>%
  ungroup() %>%  na.omit()

psem_co2_2 <- psem(
  mod1 <- lmer(swc ~ prep_2mt+hand_p+clay+ (1|area), data=df_psem2),
  mod2 <- lmer(C ~  swc_mean+hand_p+clay+ (1|area), data=df_psem2),
  mod3 <- lmer(ts ~  swc+prep_2mt + (1|area), data=df_psem2),
 # mod4 <- lmer(clay ~ hand_p + swc_mean + (1|area), data=df_psem), #I excluded clay because it is completely explained by the random element. Therefore, it is an exogenous variable.
  mod5 <- lmer(fco2 ~ swc*ts+C+clay + (swc|area/point), data=df_psem2),
  swc_mean %~~% swc,
  data=df_psem2)


summary(psem_co2_2)

plot(psem_co2_2)
```

```{r}
psem_ch4_2 <- psem(
  mod1 <- lmer(swc ~ prep_2mt+hand_p+clay+ (1|area), data=df_psem2),
  mod2 <- lmer(C ~  swc_mean+hand_p+clay+ (1|area), data=df_psem2),
  mod3 <- lmer(ts ~  swc+prep_2mt + (1|area), data=df_psem2),
 # mod4 <- lmer(clay ~ hand_p + swc_mean + (1|area), data=df_psem), #I excluded clay because it is completely explained by the random element. Therefore, it is an exogenous variable.
  mod5 <- lmer(fch4 ~ prep_2mt+swc*ts+C+clay + (1|area/point), data=df_psem2),
  swc_mean %~~% swc,
  data=df_psem2)


summary(psem_ch4_2)

plot(psem_ch4_2)
```


## Ploting

```{r}
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

plot_SEMCO2 <- grViz("
digraph SEM_Refined {
  # Layout settings to create space between nodes.
  graph [layout = dot, rankdir = TB, nodesep = 0.8, ranksep = 1.0]

  # Nodes with thin borders and smaller font.
  node [shape = box, style = 'rounded', fontname = 'Arial', 
        fontsize = 12, penwidth = 1]
  
   HAND  [label = 'HAND']
  PREP  [label = 'Precipitation']
  SWC_M [label = 'Mean SWC']
  CLAY  [label = 'Clay']
  C     [label = <Soil Carbon<br/><font point-size='8'><i>R²=0.87</i></font>>]
  SWC   [label = <Instant SWC<br/><font point-size='8'><i>R²=0.68</i></font>>]
  TS    [label = <Soil Temp<br/><font point-size='8'><i>R²=0.39</i></font>>]
  INT   [label = <Interaction<br/>SWC:TS>]
  
   FCO2  [label = <Flux CO<sub>2</sub><br/><font point-size='8'><i>R²=0.77</i></font>>, penwidth = 2]

  
  # The giants
  SWC -> FCO2 [penwidth = 5.5, color = '#e34a33', label = '-2.47']
  INT -> FCO2 [penwidth = 6.0, color = '#2c7fb8', label = '2.54']

  # Remaining (Scale from 0.5 to 2.5)
  SWC_M -> C [penwidth = 2.5, color = '#2c7fb8', label = '0.64']
  PREP -> SWC [penwidth = 2.2, color = '#2c7fb8', label = '0.59']
  C -> FCO2   [penwidth = 2.0, color = '#2c7fb8', label = '0.58']
  HAND -> SWC [penwidth = 1.8, color = '#e34a33', label = '-0.45']
  PREP -> TS  [penwidth = 1.4, color = '#e34a33', label = '-0.35']
  SWC -> TS   [penwidth = 1.2, color = '#e34a33', label = '-0.32']
  HAND -> C   [penwidth = 1.2, color = '#e34a33', label = '-0.32']
  TS -> FCO2  [penwidth = 1.0, color = '#e34a33', label = '-0.29']
  CLAY -> SWC [penwidth = 0.7, color = '#2c7fb8', label = '0.17']
  CLAY -> C   [penwidth = 0.5, color = '#2c7fb8', label = '0.14']
  
  # Non significant
  CLAY -> FCO2 [penwidth = 0.5, color = '#bdbdbd', style = dashed, label = 'ns']

  # Correlation
  SWC_M -> SWC [dir = both, style = dotted, color = '#bdbdbd', arrowsize = 0.5, label = '0.45']
}
")
```

```{r}
plot_SEMCH4 <- grViz("
digraph SEM_CH4 {
  graph [layout = dot, rankdir = TB, nodesep = 0.8, ranksep = 1.0]

  node [shape = box, style = 'rounded', fontname = 'Arial', 
        fontsize = 12, penwidth = 1]
  
  # Variables
  HAND  [label = 'HAND']
  PREP  [label = 'Precipitation']
  SWC_M [label = 'Mean SWC']
  CLAY  [label = 'Clay']
  C     [label = <Soil Carbon<br/><font point-size='8'><i>R²=0.87</i></font>>]
  SWC   [label = <Instant SWC<br/><font point-size='8'><i>R²=0.68</i></font>>]
  TS    [label = <Soil Temp<br/><font point-size='8'><i>R²=0.39</i></font>>]
  
  # Target: CH4 Flux
  FCH4  [label = <Flux CH<sub>4</sub><br/><font point-size='8'><i>R²=0.45</i></font>>, penwidth = 2]

  # --- RELATIONA ---

  # the giant (Std.Est = 1.02)
  SWC -> FCH4 [penwidth = 6.0, color = '#2c7fb8', label = '1.02']

  # remaining
  PREP -> FCH4 [penwidth = 2.5, color = '#2c7fb8', label = '0.36']
  TS -> FCH4   [penwidth = 1.5, color = '#2c7fb8', label = '0.17']

  # non-significant (Dashed/Gray)
  C -> FCH4    [penwidth = 1.0, color = '#bdbdbd', style = dashed, label = 'ns']
  CLAY -> FCH4 [penwidth = 0.5, color = '#bdbdbd', style = dashed, label = 'ns']

  # structure (Repetindo os drivers do solo/clima)
  SWC_M -> C   [penwidth = 2.5, color = '#2c7fb8', label = '0.64']
  PREP -> SWC  [penwidth = 2.2, color = '#2c7fb8', label = '0.59']
  HAND -> SWC  [penwidth = 1.8, color = '#e34a33', label = '-0.45']
  PREP -> TS   [penwidth = 1.4, color = '#e34a33', label = '-0.35']
  SWC -> TS    [penwidth = 1.2, color = '#e34a33', label = '-0.32']
  HAND -> C    [penwidth = 1.2, color = '#e34a33', label = '-0.32']
  CLAY -> SWC  [penwidth = 0.7, color = '#2c7fb8', label = '0.17']
  CLAY -> C    [penwidth = 0.5, color = '#2c7fb8', label = '0.14']
  
  # correlation
  SWC_M -> SWC [dir = both, style = dotted, color = '#bdbdbd', arrowsize = 0.5, label = '0.45']
}
")
```

```{r}
export_svg(plot_SEMCO2) %>%
  charToRaw() %>%
  rsvg_png("Plots/PSEM_co2.png", width = 1890, height = 1231)


export_svg(plot_SEMCH4) %>%
  charToRaw() %>%
  rsvg_png("Plots/PSEM_ch4.png", width = 1890, height = 1231)
```

