---
title: "TOMSTMoisture"
author: "Larissa Verona"
date: "2026-04-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data analyses to find moisture patters with TOMST data

## Data and Packages

```{r}
rm(list = ls())
library(tidyverse)
library(ggthemes)
library(feasts)
library(tsibble)
library(purrr)
library(lubridate)
library(ggpubr)
library(emmeans)
library(multcomp)
```

```{r}
# Check if data exists
data_path <- "Data/Moisture/tms_data.csv"

if (!file.exists(data_path)) {
  stop(
    "File 'tms_data.csv' not found.\n",
    "Please download it from: https://drive.google.com/file/d/1--GgRlHxu_I2oiT_HL_feh3D9u9p3fWD/view?usp=drive_link\n",
    "and place it in 'Data/Moisture/'."
  )
}
```

The data used in this analysis are generated using the script
TOMSTMoistureProcessing.Rmd included in this project. Due to their large
size, the data are not stored in the GitHub repository. The processed
dataset is available at:
<https://drive.google.com/file/d/1--GgRlHxu_I2oiT_HL_feh3D9u9p3fWD/view?usp=drive_link>
. Please download the file and place it in the Data/Moisture/ directory.

```{r}
tms_data <- read.csv2("Data/Moisture/tms_data.csv", sep=",") %>%
  mutate(value=as.numeric(value)) %>%
  pivot_wider(names_from = sensor_name,
              values_from = value,
              id_cols = c(locality_id, datetime)) %>%
  separate(locality_id, 
           into=c("area", "point"), 
           sep= "_") %>%
  mutate( point=toupper(point),
         area=tolower(area),
         area=case_when(area =="7q"~"seteq",
         TRUE ~ as.character(area)),
         area=as.factor(area))

tms_data <- tms_data %>% 
  mutate(datetime = if_else(
    stringr::str_detect(datetime, " "),          # já tem hora?
    datetime,
    paste0(datetime, " 00:00:00")          # adiciona hora
  )) %>% 
  mutate(datetime = ymd_hms(datetime))    # converte para POSIXct
  
tms_data <- tms_data %>%
  mutate(
    point = case_when(
      area == "ing" & point == "P10A" ~ "P10",
      area == "ing" & point == "P10B" ~ "P10",
      area == "ing" & point == "P2C"  ~ "P2",
      area == "ing" & point == "P2A"  ~ "P2",
      area == "ing" & point == "P4A"  ~ "P4",
      area == "ing" & point == "P6A"  ~ "P6",
      area == "ing" & point == "P8A"  ~ "P8",
      TRUE ~ point
    )
  ) %>%
  filter(!(area == "ing" & point %in% c("P2B", "P4B", "P6B", "P8B")))
  
tms_data <- tms_data %>% mutate(datetime=datetime-hours(3),
                                point=as.factor(point)) #ajustando de UTC para GMT-3
str(tms_data)
tms_data <- tms_data %>% filter(off_soil==0) 
```

Diary data

```{r}
tms_day <- tms_data %>%  mutate(day = lubridate::floor_date(datetime, unit="day")) %>%
    group_by(area, point, day) %>%
  summarise(TMS_moist=mean(TMS_moist), VWC_moisture=mean(VWC_moisture))
```

Week data

```{r}
tms_week <- tms_data %>%
  mutate(week = lubridate::floor_date(datetime, unit = "week")) %>%
  group_by(area, point, week) %>%
  summarise(
    TMS_moist = mean(TMS_moist, na.rm = TRUE),
    VWC_moisture = mean(VWC_moisture, na.rm = TRUE),
    .groups = "drop" 
  )
```

## Ploting

```{r}
##ing
tms_day %>% filter(area=="ing") %>%
mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10"))) %>% 
  ggplot(aes(x=day, y=VWC_moisture)) +
  geom_line(color="darkblue")+
  facet_wrap(~point, ncol=1) +
  xlab("")+
  ylab("Volumetric Water Content (%)")+
  theme_clean()+
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12)) +
  ggtitle("ING")



ggsave("Plots/moist_ing.png", unit="cm", width = 10, height = 15)  


##vvt
tms_data %>% filter(area=="vvt") %>%
  mutate(dia = lubridate::floor_date(datetime, unit="day")) %>%
    group_by(point, dia) %>%
  summarise(VWC_moisture=mean(VWC_moisture)) %>%
mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10"))) %>%
  ggplot(aes(x=dia, y=VWC_moisture)) +
  geom_line(color="darkblue")+
  facet_wrap(~point, ncol=1) +
  xlab("")+
  ylab("Volumetric Water Content (%)")+
  theme_clean()+
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12))  +
  ggtitle("VVT")


ggsave("Plots/moist_vvt.png", unit="cm", width = 10, height = 15)  

##cr
tms_data %>% filter(area=="cr") %>%
  mutate(dia = lubridate::floor_date(datetime, unit="day")) %>%
    group_by(point, dia) %>%
  summarise(VWC_moisture=mean(VWC_moisture)) %>%
 mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10"))) %>%
  ggplot(aes(x=dia, y=VWC_moisture)) +
  geom_line(color="darkblue")+
  facet_wrap(~point, ncol=1) +
  xlab("")+
  ylab("Volumetric Water Content (%)")+
  theme_clean()+
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12)) +
    ggtitle("CR")



ggsave("Plots/moist_cr.png", unit="cm", width = 10, height = 15)

##7Q
tms_data %>% filter(area=="seteq") %>%
  mutate(dia = lubridate::floor_date(datetime, unit="day")) %>%
    group_by(point, dia) %>%
  summarise(VWC_moisture=mean(VWC_moisture)) %>%
mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10"))) %>%
  ggplot(aes(x=dia, y=VWC_moisture)) +
  geom_line(color="darkblue")+
  facet_wrap(~point, ncol=1) +
  xlab("")+
  ylab("Volumetric Water Content (%)")+
  theme_clean()+
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12)) +
  ggtitle("7Q")



ggsave("Plots/moist_7Q.png", unit="cm", width = 10, height = 15) 

##BBT
tms_data %>% filter(area=="bbt") %>%
  mutate(dia = lubridate::floor_date(datetime, unit="day")) %>%
    group_by(point, dia) %>%
  summarise(VWC_moisture=mean(VWC_moisture)) %>%
mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10"))) %>%
  ggplot(aes(x=dia, y=VWC_moisture)) +
  geom_line(color="darkblue")+
  facet_wrap(~point, ncol=1) +
  xlab("")+
  ylab("Volumetric Water Content (%)")+
  theme_clean()+
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12))+
  ggtitle("BBT")


ggsave("Plots/moist_bbt.png", unit="cm", width = 10, height = 15) 

```

Ploting a heatmap

To use a better approach, I'll try to add the HAND values instead of
points

HAND

```{r}
  hand <- read.csv("Data/Hand/Flux_points_HAND.csv", sep=";") %>% 
  dplyr::select("area", "point", "hand") %>% 
  group_by(area) %>%
  mutate(hand_p = (hand-min(hand))/(max(hand)-min(hand)))

tms_data_hm <-  left_join(tms_day, hand, by=c("area", "point")) %>% mutate(area=factor(area),
       point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6", "P7", "P8", "P9", "P10"))) %>%
  mutate(point = reorder(point, hand)) 


```

HeatMap

```{r}

library(scales)

labels_areas <- c(
  "bbt" = "BBT",
  "ing" = "ING",
  "seteq" = "7Q",
  "cr" = "CR",
  "vvt" = "VVT"
)
tms_data_hm %>% ungroup()%>% 
 
  mutate(point_label = paste0(point, " (", round(hand, 2), "m)"),
point_label = reorder(point_label, hand)) %>% 
ggplot(aes(x = day, y = point_label, fill = TMS_moist)) +
  geom_tile() +
  scale_fill_viridis_c(option = "mako", name = "Moisture Signal") +
  scale_x_date(labels = label_date("%b %Y", locale = "en")) +
  labs(
    x = "",
    y = "Point (HAND elevation)"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.position = "bottom",
    panel.grid = element_blank() 
  ) +
  facet_wrap(~area, scales = "free", labeller = as_labeller(labels_areas))
```

## Obteining Data from TS

Exploring

```{r}
tms_day %>% 
  filter(area == "bbt", point == "P10") %>% 
  arrange(day) %>% 
  pull(TMS_moist) %>% 
 diff()  %>%   # A mágica acontece aqui: calcula a diferença entre t e t-1
   acf(main = "Temporal Correlation for Differences - BBT P10 ")

ggsave("Plots/ACF_BBT_P10.png", unit="cm", width = 10, height = 15)

tms_day %>% 
  filter(area == "bbt", point == "P3") %>% 
  arrange(day) %>% 
  pull(TMS_moist) %>% 
 diff()  %>%     # A mágica acontece aqui: calcula a diferença entre t e t-1
  acf(main = "Temporal Correlation for Differences - BBT P3 ")

ggsave("Plots/ACF_BBT_P3.png", unit="cm", width = 10, height = 15)
```

### Autocorrelation and entropy

Vou obter: - Autocorrelação temporal com lag1 a lag 10 para a série
semanal de umidade --\> corresponde a correlação dos valores de cada
grupo de 10 semanas. - Autocorrelação temporal da serie de diferenciais
diárias para lag 1 a 4 --\> corresponde a tendencias de subida ou
descida - Entropia para série de 15 min - Entropia para série diária

Data

```{r}
ts_15min <- tms_data %>% 
  dplyr::select(area, point, datetime, TMS_moist, VWC_moisture) %>% 
  as_tsibble(key = c(area, point), index = datetime)

ts_daily <- tms_day %>% 
  ungroup() %>%
  mutate(day = as.Date(day)) %>% 
  as_tsibble(key = c(area, point), index = day)

ts_weekly <- tms_week %>% 
  ungroup() %>%
  mutate(week = yearweek(week)) %>% 
  as_tsibble(key = c(area, point), index = week)

```

Calculating ACFs and ACFs diff

```{r}

# 1. Function to lags from differential series (daily serie)
diff_lags_feat <- function(x) {
  dx <- diff(x)
  # Calculamos o ACF até o lag 4
  res <- acf(dx, plot = FALSE, na.action = na.pass, lag.max = 4)$acf
  # O índice [1] é o lag 0, então pegamos [2:5]
  return(list(diff_acf1 = res[2], diff_acf2 = res[3], 
              diff_acf3 = res[4], diff_acf4 = res[5]))
}

# 2. Extracting daily metrics (Diferential lags)
feat_daily <- ts_daily %>%
  features(TMS_moist, diff_lags_feat)

# 3. Extracting weekly metrics (Lag 1 e Lag 10)
feat_weekly <- ts_weekly %>%
  features(TMS_moist, feat_acf) %>%
  dplyr::select(area, point, acf1, acf10)

# 4. Extracting entropy
feat_entropy15 <- ts_15min %>%
  features(TMS_moist, list(entropy = feat_spectral)) %>%
  rename(entropy_15 = entropy_spectral_entropy)

feat_entropyday <- ts_daily %>%
  features(TMS_moist, list(entropy = feat_spectral))  %>%
  rename(entropy_day = entropy_spectral_entropy)

feat_entropywk <- ts_weekly %>%
  features(TMS_moist, list(entropy = feat_spectral))  %>%
  rename(entropy_week = entropy_spectral_entropy)

TS_features <- feat_entropy15 %>%
  left_join(feat_entropyday, by = c("area", "point")) %>% 
  left_join(feat_entropywk, by = c("area", "point")) %>%
  left_join(feat_weekly, by = c("area", "point")) %>%
  left_join(feat_daily, by = c("area", "point")) %>%
  mutate(across(starts_with("diff_acf"), as.numeric))
```

Calculating ACFs and ACFs diff for wet and dry season

```{r}

#1. Classifying the season
ts_daily <- ts_daily %>% 
  mutate(season = case_when(
    month(day) %in% c(7, 8, 9,10) ~ "dry",
    month(day) %in% c(12, 1, 2, 3) ~ "wet",
    TRUE ~ "transition"
  ))

ts_weekly <- ts_weekly %>% 
  mutate(season = case_when(
    month(week) %in% c(7, 8, 9,10) ~ "dry",
    month(week) %in% c(12, 1, 2, 3) ~ "wet",
    TRUE ~ "transition"
  ))

ts_15min <- ts_15min %>% 
  mutate(season = case_when(
    month(datetime) %in% c(7, 8, 9,10) ~ "dry",
    month(datetime) %in% c(1, 2, 3) ~ "wet",
    TRUE ~ "transition"
  ))

#2. Extracting daily metrics (Diferential lags)
feat_daily_wt <- ts_daily %>% 
  filter(season=="wet") %>%
  features(TMS_moist, diff_lags_feat) %>% rename(
    diff_acf1_wt = diff_acf1,
    diff_acf2_wt = diff_acf2,
    diff_acf3_wt = diff_acf3,
    diff_acf4_wt = diff_acf4
  )

feat_daily_dr <- ts_daily %>% 
  filter(season=="dry") %>%
  features(TMS_moist, diff_lags_feat) %>%
 rename(
    diff_acf1_dr = diff_acf1,
    diff_acf2_dr = diff_acf2,
    diff_acf3_dr = diff_acf3,
    diff_acf4_dr = diff_acf4
  )

# 3. Extracting weekly metrics (Lag 1 e Lag 10)
feat_weekly_wt <- ts_weekly %>% 
  filter(season=="wet") %>%
  features(TMS_moist, feat_acf) %>%
  dplyr::select(area, point, acf1, acf10) %>%
  rename(
    acf1_wt = acf1,
    acf10_wt = acf10)

feat_weekly_dr <- ts_weekly %>% 
  filter(season=="dry") %>%
  features(TMS_moist, feat_acf) %>%
  dplyr::select(area, point, acf1, acf10) %>%
   rename(
    acf1_dr = acf1,
    acf10_dr = acf10)


# 4. Extracting entropy
feat_entropy15_wt <- ts_15min %>% 
  filter(season=="wet") %>%
  features(TMS_moist, list(entropy = feat_spectral)) %>%
  rename(entropy_15_wt = entropy_spectral_entropy)

feat_entropy15_dr <- ts_15min %>% 
  filter(season=="dry") %>%
  features(TMS_moist, list(entropy = feat_spectral)) %>%
  rename(entropy_15_dr = entropy_spectral_entropy)

feat_entropyday_wt <- ts_daily %>% 
  filter(season=="wet") %>%
  features(TMS_moist, list(entropy = feat_spectral))  %>%
  rename(entropy_day_wt = entropy_spectral_entropy)

feat_entropyday_dr <- ts_daily %>% 
  filter(season=="dry") %>%
  features(TMS_moist, list(entropy = feat_spectral))  %>%
  rename(entropy_day_dr = entropy_spectral_entropy)

feat_entropywk_wt <- ts_weekly %>% 
  filter(season=="wet") %>%
  features(TMS_moist, list(entropy = feat_spectral))  %>%
  rename(entropy_week_wt = entropy_spectral_entropy)

feat_entropywk_dr <- ts_weekly %>% 
  filter(season=="dry") %>%
  features(TMS_moist, list(entropy = feat_spectral))  %>%
  rename(entropy_week_dr = entropy_spectral_entropy)


TS_features_season <- feat_entropy15_wt %>%
  left_join(feat_entropy15_dr, by = c("area", "point")) %>%
  left_join(feat_entropyday_wt, by = c("area", "point")) %>%
  left_join(feat_entropyday_dr, by = c("area", "point")) %>% 
  left_join(feat_entropywk_wt, by = c("area", "point")) %>%
  left_join(feat_entropywk_dr, by = c("area", "point")) %>%
  left_join(feat_weekly_wt, by = c("area", "point")) %>%
  left_join(feat_weekly_dr, by = c("area", "point")) %>%
  left_join(feat_daily_wt, by = c("area", "point")) %>%
  left_join(feat_daily_dr, by = c("area", "point")) %>%
  mutate(across(starts_with("diff_acf"), as.numeric))
```

Weak patterns, apparently

```{r}
TS_features_season %>% mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10"))) %>%
  ggplot(aes(x=point, y=diff_acf4_dr)) +
  geom_point()+
  geom_hline(yintercept = 0)+
  facet_wrap(~area)
  
```

### Mean, SD and Saturation lenght

```{r}
# Function to count continuous days with saturation (Run Length Encoding)
max_continuous_sat <- function(is_sat) {
  if(!any(is_sat)) return(0)
  runs <- rle(is_sat)
  max(runs$lengths[runs$values == TRUE], na.rm = TRUE)
}

metrics_base <- tms_day %>%
  group_by(area, point) %>%
  summarise(
    max = max(TMS_moist, na.rm = TRUE),
    min = min(TMS_moist, na.rm = TRUE),
    mean = mean(TMS_moist, na.rm = TRUE),
    sd = sd(TMS_moist, na.rm = TRUE))
    
    # Saturation (saturated if moisture > 95% from point max)
metrics_base2 <- tms_day %>%
  group_by(area, point) %>%
     mutate(is_saturated = TMS_moist >= (max(TMS_moist) * 0.95)) %>% summarise(days_sat_total = sum(is_saturated, na.rm = TRUE),
    days_sat_cont  = max_continuous_sat(is_saturated),
    .groups = "drop"
  )



TS_features <- TS_features_season %>%
  left_join(TS_features, by = c("area", "point")) %>%
  left_join(metrics_base, by = c("area", "point")) %>%
  left_join(metrics_base2, by = c("area", "point"))
```

### Wet season stability

```{r}
stability_wet_fixed <- tms_day %>%
  mutate(mes = month(day)) %>%
  # filtring moisture season
  filter(mes %in% c(12, 1, 2, 3)) %>% 
  group_by(area, point) %>%
  summarise(
    wet_mean = mean(TMS_moist, na.rm = TRUE),
    wet_sd   = sd(TMS_moist, na.rm = TRUE),
    # low CV = High stability
    wet_cv   = wet_sd / wet_mean, 
    .groups = "drop"
  )

TS_features <- left_join(TS_features, stability_wet_fixed, by=c("area", "point"))
```

### Transition slope

Obteining the transition segments

```{r}
#ploting
tms_day %>% 
  filter(area == "bbt") %>%
  ggplot(aes(x = day, y = TMS_moist)) +
  geom_line(alpha = 0.4, color = "grey50") +
  geom_smooth(method = "loess", span = 0.5, color = "red", se = FALSE) +
  facet_wrap(~point, scales = "free_y") +
  theme_clean() +
  labs(title = "Tendência de Longo Prazo via LOESS (Área BBT)",
       x = "", y = "TMS Moisture")


#find a loess function
tms_smooth <- tms_day %>%
  group_by(area, point) %>%
  mutate(x_num = as.numeric(day)) %>% 
  group_modify(~ {
    # LOESS model fited to actual group
    # span says how strict the model is
    fit <- loess(TMS_moist ~ x_num, data = .x, span = 0.6)
    # Adiciona os valores preditos  ao dataframe
    .x %>% mutate(TMS_trend = predict(fit))
  }) %>%
  ungroup() %>%
  dplyr::select(-x_num)

#ploting
tms_smooth %>%
  filter(area == "ing") %>%
  ggplot(aes(x = day)) +
  geom_line(aes(y = TMS_moist), alpha = 0.6) + 
  geom_line(aes(y = TMS_trend), color = "red", size = 1) + 
  labs(title = "Série Temporal Original vs. Dados LOESS Salvos") +
  facet_wrap(~point)

#identifing transition
  find_recession_segment <- function(df_group) {
  # 1. Finding the min point (drought peak)
  idx_min <- which.min(df_group$TMS_trend)
  
  # 2. scan data before min
  df_pre_min <- df_group[1:idx_min, ]
  
  # 3. Derivative (daily variation)
  # We look for last derivative >=0 (start transition)
  derivada <- diff(df_pre_min$TMS_trend)
  
  # Finding the last point where derivative was >= 0 before continous slope
  idx_start <- max(which(derivada >= 0)) + 1
  
  # If we do not find it, we use the series begging
  if(is.infinite(idx_start)) idx_start <- 1
  
  # Getting the data from this point
  df_group[idx_start:idx_min, ]
  }
  
df_seg <- tms_smooth %>%
  group_by(area, point) %>%
  group_modify(~ find_recession_segment(.x)) %>%
  ungroup()

# 3. Fast validation: How long was the slope in each point?
seg_summary <- df_seg %>%
  group_by(area, point) %>%
  summarise(
    transition_begin = min(day),
    transition_end = max(day),
    transition_duration = n(),
    trans_total_lost = first(TMS_trend) - last(TMS_trend),
    .groups = "drop"
  )


View(seg_summary)
write.csv(seg_summary, "Data/Moisture/transition_data.csv")
```

```{r}
tms_smooth <- tms_smooth %>% mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10")))

df_seg <- df_seg %>% mutate(point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6","P7", "P8", "P9", "P10")))

ggplot() +
  geom_line(data = tms_smooth %>% filter(area=="vvt"), aes(x = day, y = TMS_moist), color = "grey85") +
  geom_line(data = df_seg %>% filter(area=="vvt"), aes(x = day, y = TMS_trend), color = "blue", size = 1) +
  facet_wrap(~point) +
  theme_minimal() +
  labs(title = "VVT")
ggsave("Plots/Prep+trans_VVT.png")

TS_features <- left_join(TS_features, seg_summary, by=c("area", "point"))

TS_features <- TS_features %>% mutate(transition_delay = as.Date(transition_begin) - as.Date("2024-11-02")) %>% mutate(transition_delay = as.numeric(transition_delay)) %>% dplyr::select(-transition_begin, -transition_end)
```

Obteining the variation per point

```{r}
# Calculating the daily variation
df_diff <- df_seg %>%
  group_by(area, point) %>%
  arrange(day) %>%
  mutate(
    slope_diario = TMS_trend - lag(TMS_trend)
  ) %>%
  ungroup()

df_diff <- df_diff %>% filter(!is.na(slope_diario))


df_diff %>%
  filter(area == "cr") %>%
  ggplot(aes(x = day, y = slope_diario)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  # Slope line
  geom_line(color = "firebrick", size = 1) +
  geom_area(fill = "firebrick", alpha = 0.2) +
  facet_wrap(~point) +
  theme_minimal() +
  labs(
    title = "Instant Drying Speed (Derived from LOESS)",
    x = "", 
    y = "Daily variation (unit/day)"
  )
```

Obteining transion metrics

```{r}
transition_metrics <- df_diff %>%
  group_by(area, point) %>%
  summarise(
    # 1. Linear Slope
    # (Final_value - Initial_value) / Delta_time
    transition_slope = (last(TMS_trend) - first(TMS_trend)) / n(),
    
    # 2. Loss max speed(Derivative peak)
    transition_max_slope = min(slope_diario, na.rm = TRUE),
    
    # 3. Drying stability (derivative SD)
    transition_stability = sd(slope_diario, na.rm = TRUE),
    
    .groups = "drop"
  )

TS_features <- left_join(TS_features, transition_metrics, by=c("area", "point"))
```

### WorldClim Variables

Moisture in the wettest and driest month

```{r}
wc_var <- tms_day %>%
    mutate(month = lubridate::floor_date(day, unit = "month")) %>%
  group_by(area, point, month) %>%
  summarise(
    TMS_moist_mes = mean(TMS_moist, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(area, point) %>%
  summarise(
    wettest_month = max(TMS_moist_mes, na.rm = TRUE),
    driest_month = min(TMS_moist_mes, na.rm = TRUE),
    .groups = "drop"
  )
```

Moisture in the wettest and driest quarter

```{r}
wc_var2 <- tms_day %>%
    mutate(quarter = lubridate::floor_date(day, unit = "quarter")) %>%
  group_by(area, point, quarter) %>%
  summarise(
    TMS_moist_qt = mean(TMS_moist, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(area, point) %>%
  summarise(
    wettest_quarter = max(TMS_moist_qt, na.rm = TRUE),
    driest_quarter = min(TMS_moist_qt, na.rm = TRUE),
    .groups = "drop"
  )

TS_features <- left_join(TS_features, wc_var, by=c("area", "point"))
TS_features <- left_join(TS_features, wc_var2, by=c("area", "point"))
```

## Soil Data

```{r}
soil <- read.csv("Data/Soils/Soil.csv", sep=";")

df <- left_join(TS_features, soil %>% dplyr::select(-sand, -silt, -N), by=c("area", "point"))
str(df)
```

## HAND data

```{r}
hand <- read.csv("Data/Hand/Flux_points_HAND.csv", sep=";") %>% 
  dplyr::select("area", "point", "hand") %>% 
  group_by(area) %>%
  mutate(hand_p = (hand-min(hand))/(max(hand)-min(hand)))

df <-  left_join(df, hand, by=c("area", "point")) %>% mutate(area=factor(area),
       point=factor(point, levels=c("P1", "P2", "P3", "P4", "P5", "P6", "P7", "P8", "P9", "P10"))) %>% dplyr::select(-hand)
```

## PCA and Clustering

Scaling

```{r}
df_main <- df[, c(9,10,11,12, 16,20,30,31,32,33,34,35,36,38,39,40,41,42,43,44,45,46)]

df_scaled <- df_main %>% 
  mutate(across(where(is.numeric), ~ as.vector(scale(.))))

df_numeric <- df_scaled %>% dplyr::select(where(is.numeric))
```


```{r}
library(ggcorrplot)  
  
  corr_matrix <- cor(df_numeric %>% na.omit())
  
  # Criar o gráfico
  ggcorrplot(corr_matrix, 
                  hc.order = TRUE, 
                  type = "lower",
                  lab = F, 
             show.diag = FALSE,
                  colors = c("#6D9EC1", "white", "#E46726"))


```

PCA

```{r}
library(factoextra)


df_numeric <- df_numeric %>%
  mutate(across(everything(), ~replace_na(., median(., na.rm = TRUE))))
pca_res <- prcomp(df_numeric, scale. = TRUE)
fviz_eig(pca_res, addlabels = TRUE, ylim = c(0, 70))

df_pca_scores <- pca_res$x[, 1:3]
```

```{r}
fviz_pca_biplot(pca_res, 
                geom.ind = "point",
                fill.ind = df$area,
                pointshape = 21, pointsize = 2.5,
                palette = "jco",
                col.var = "black",
                alpha.var = 0.5,
                repel = TRUE, 
                legend.title = "Áreas") +
  theme_minimal() 
```

K-means

```{r}
set.seed(123) 

# K-means 1: Raw Data
km_bruto <- kmeans(df_numeric, centers = 3, nstart = 25)

# K-means 2: PCA Data
km_pca <- kmeans(df_pca_scores, centers = 3, nstart = 25)

df_final <- df %>%
  mutate(cluster_bruto = as.factor(km_bruto$cluster),
         cluster_pca = as.factor(km_pca$cluster))
```

```{r}
library(patchwork) 

node_labels <- paste0(df$area, "_", df$point)

rownames(df_numeric) <- node_labels
rownames(df_pca_scores) <- node_labels

# K-means 1
p1 <- fviz_cluster(km_bruto, 
                   data = df_numeric,
                   geom = c("point", "text"), 
                   repel = TRUE,              
                   pointsize = 2,
                   labelsize = 8,
                   main = "Clusters: Raw Data") +
  theme_minimal()

# K-means 2
p2 <- fviz_cluster(km_pca, 
                   data = df_pca_scores,
                   geom = c("point", "text"), 
                   repel = TRUE,
                   pointsize = 2,
                   labelsize = 8,
                   main = "Clusters: PCA Scores") +
  theme_minimal()

p1 + p2

ggsave("Plots/clusters.png", unit="cm", width = 25, height = 10)

```

## Soil Flux

```{r}

# 1. List all csv
files <- list.files(
  path = "Data/FluxProcessed",
  pattern = "^(vvt|7q|bbt|cr|ing).+\\.csv$",
  full.names = TRUE
)

# 2. Function to deal with different formats
read_allfiles <- function(arq) {
  
  # Detectar o separador
  sep_df <- if(grepl(";", readLines(arq, n = 1))) ";" else ","
  
  # 1. Pegar apenas o nome do arquivo (sem o caminho das pastas)
  # 2. Remover a extensão .csv
  file_name <- str_remove(basename(arq), "\\.csv")
  
  # 3. Agora o split vai funcionar como esperado
  meta <- str_split(file_name, "_")[[1]]
  
  df_temp <- read.csv(arq, sep = sep_df, skip = 1) %>%
    slice(-1) %>%
    # Agora meta[1] será "vvt" e meta[2] será "set24"
    mutate(area = meta[1], month = meta[2])
  
  # Lógica de nomes de coluna
  if ("DATE_TIME.initial_value" %in% names(df_temp)) {
    df_temp <- df_temp %>% rename(datetime = DATE_TIME.initial_value)
  }
  
  # Formatos de data
  df_temp <- df_temp %>%
    mutate(datetime = parse_date_time(datetime, orders = c("dmy HM", "ymd HMS")))
  
  return(df_temp)
}

# 3. Processing
df_flux <- files %>% 
  map_dfr(read_allfiles) %>%
  
  mutate(across(c(3, 4, 7:14), as.numeric),
         across(c(2, 16, 17), as.factor)) %>%
  rename(point = COMMENT, 
         swc = SWC_2.mean, 
         ts = TS_2.mean, 
         fch4 = FCH4_DRY, 
         fco2 = FCO2_DRY,
         lat = LATITUDE,
         long = LONGITUDE) %>%
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  mutate(area=case_when(area =="7q"~"seteq",
         TRUE ~ as.character(area)))

str(df_flux)
rm(files)
```

## Correlating Fluxes to Clusters

```{r}
df_fluxTST <- df_flux %>% dplyr::select(c("area", "point", "month", "fco2", "fch4", "swc")) %>% right_join(df_final, by=c("area", "point"))

df_fluxTST %>% 
  ggplot(aes(x=cluster_bruto, y=fco2)) +
  geom_boxplot()+
  geom_jitter()+
  stat_compare_means()+
  facet_wrap(~month)

df_fluxTST %>% 
  ggplot(aes(x=cluster_bruto, y=fch4)) +
  geom_boxplot()+
  geom_jitter()+
  stat_compare_means()+
  facet_wrap(~month)

df_cluster_means <- df_fluxTST %>%
  group_by(cluster_bruto) %>%
  summarise(CH4_mean = mean(fch4), na.rm = TRUE,
            CO2_mean = mean(fco2), na.rm = TRUE)

df_fluxTST %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>% 
  mutate(point2 = paste(area, point)) %>%
  ggplot(aes(x = month, y = fco2, color = swc, group = point2)) + 
  geom_point(size = 1.2, alpha=0.8) +
  geom_line(alpha = 0.5) + # Adiciona as linhas conectando os pontos +
  theme_clean()+
  geom_smooth(aes(group = 1), method = "loess", color = "red", linetype = "dashed", se = T) +
  geom_hline(data = df_cluster_means, 
             aes(yintercept = CO2_mean), 
             color = "black", 
             linewidth = 0.5) +
  facet_wrap(~cluster_bruto) +
  ggtitle("Clusters Annual Emissions") +
  ylab(expression(log~CO[2]~(mu*mol ~ m^{-2} * s^{-1})))+
  xlab("") +
  labs(color="SWC (%)")


ggsave("Plots/Emissions_clustersco2.png", unit="cm", width = 25, height = 15)



df_fluxTST %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>% 
  mutate(point2 = paste(area, point)) %>%
  ggplot(aes(x = month, y = fch4, color = swc, group = point2)) + 
  theme_clean()+
  geom_point (size = 1.2, alpha=0.8) +
  geom_line(alpha = 0.5) + # Adiciona as linhas conectando os pontos +
  geom_smooth(aes(group = 1), method = "loess", color = "red", linetype = "dashed", se = T) +
  geom_hline(data = df_cluster_means, 
             aes(yintercept = CH4_mean), 
             color = "black", 
             linewidth = 0.5) +
  geom_hline(aes(yintercept = 0),
             color = "black", 
             linewidth = 0.5,
             linetype = "dashed"
          ) +
  facet_wrap(~cluster_bruto) +
     ggtitle("Clusters Annual Emissions") +
  ylab(expression(log~CH[]~(nmol ~ m^{-2} * s^{-1})))+
  xlab("") +
  labs(color="SWC (%)")


ggsave("Plots/Emissions_clustersch4.png", unit="cm", width = 25, height = 15)
 
  df_fluxTST %>% 
  ggplot(aes(x=cluster_bruto, y=fch4)) +
  geom_boxplot()+
  geom_jitter()+
  stat_compare_means()+
  facet_wrap(~month) 

df_final %>% 
  ggplot(aes(x=cluster_pca, y=C)) +
  geom_boxplot()+
  geom_jitter()+
  stat_compare_means()

df_final %>% 
  ggplot(aes(x=hand, y=transition_slope)) +
  geom_point()+
  geom_smooth()+
  stat_cor()
```

Plot describing clusters

```{r}
C <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=C)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

acf <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=acf10)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

diffacf <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=diff_acf4)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

entro <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=entropy_day)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

max <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=max)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

min <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=min)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

mean <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=mean)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

sd <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=sd)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

wet_mean <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=wet_mean)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

wet_cv <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=wet_cv)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

trans_slope<- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=transition_slope)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

trans_delay<- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=transition_delay)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

trans_stability<- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=transition_stability)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

C<- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=C)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

hand_p <- df_final %>% 
  ggplot(aes(x=cluster_bruto, y=hand_p)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

hand <- df_final %>% 
  ggplot(aes(x=cluster_pca, y=hand)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

wet_month <- df_final %>% 
  ggplot(aes(x=cluster_pca, y=wettest_month)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

dry_month <- df_final %>% 
  ggplot(aes(x=cluster_pca, y=driest_month)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

wet_quarter <- df_final %>% 
  ggplot(aes(x=cluster_pca, y=wettest_quarter)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)

dry_quarter <- df_final %>% 
  ggplot(aes(x=cluster_pca, y=driest_quarter)) +
  geom_boxplot()+
  geom_jitter()+
  theme_clean()+
  stat_compare_means(label = "p.format", label.x = 1)
```

```{r}

plots <- list(
  C = C + labs(title = "Soil Organic Carbon", x = "Cluster", y = "C (%)"),
  acf = acf + labs(title = "Autocorrelation (Lag 10)", x = "Cluster", y = "ACF10"),
  diffacf = diffacf + labs(title = "First Difference ACF (Lag 4)", x = "Cluster", y = "Diff ACF4"),
  entro = entro + labs(title = "Daily Entropy", x = "Cluster", y = "Entropy"),
  max = max + labs(title = "Maximum Soil Moisture", x = "Cluster", y = "Max VWC"),
  min = min + labs(title = "Minimum Soil Moisture", x = "Cluster", y = "Min VWC"),
  mean = mean + labs(title = "Mean Soil Moisture", x = "Cluster", y = "Mean VWC"),
  sd = sd + labs(title = "Standard Deviation", x = "Cluster", y = "SD"),
  wet_mean = wet_mean + labs(title = "Wet Period Mean", x = "Cluster", y = "Wet Mean VWC"),
  wet_cv = wet_cv + labs(title = "Wet Period CV", x = "Cluster", y = "CV (%)"),
  trans_slope = trans_slope + labs(title = "Transition Slope", x = "Cluster", y = "Slope (unit/day)"),
  trans_delay = trans_delay + labs(title = "Transition Delay", x = "Cluster", y = "Days"),
  trans_stability = trans_stability + labs(title = "Transition Stability", x = "Cluster", y = "SD of Slope"),
  hand_p = hand_p + labs(title = "HAND (Topography)", x = "Cluster", y = "Height (m)")
)
```

```{r}
# Painel 1
panel_1 <- (plots$C |  plots$hand_p|plots$max) / 
           (plots$min | plots$mean | plots$wet_cv) /
           (plots$entro | plots$trans_slope | plots$trans_delay) 

panel_1


ggsave("Plots/ClusterDescription.png", unit="cm", width = 30, height = 25)
```

Testing CO2

```{r}
anova_co2 <- aov(log(fco2+1) ~ cluster_pca*month, data=df_fluxTST)
summary(anova_co2)

em_co2 <- emmeans(anova_co2, ~ cluster_pca | month)
pairs_co2 <- pairs(em_co2, adjust = "tukey")
summary(pairs_co2)
cld_co2 <- cld(em_co2, Letters = letters, adjust = "tukey")
print(cld_co2)
```

```{r}
df_letras_co2 <- as.data.frame(cld_co2) %>%
    mutate(.group = trimws(.group)) 


df_letras_co2 <- df_letras_co2 %>%
  mutate(
    erro_padrao = if("SE" %in% names(.)) SE else stderr,
    emmean = as.numeric(emmean),
    erro_padrao = as.numeric(erro_padrao)
  ) %>%
  group_by(month) %>%
  mutate(y_pos = max(emmean, na.rm = TRUE) + (2 * max(erro_padrao, na.rm = TRUE))) %>%
  ungroup()
```

```{r}
df_fluxTST %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  ggplot(aes(x = cluster_pca, y = log(fco2 + 1))) +
  
  # Boxplot e Jitter
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(alpha = 0.3, width = 0.1, color = "grey40") +
  
  # Camada de Letras (Post-hoc)
  geom_text(data = df_letras_co2, 
            aes(x = cluster_pca, y = y_pos, label = .group), 
            vjust = -0.5,           
            fontface = "bold", 
            size = 4, 
            color = "black") +
  
  # Facet por mês
  facet_wrap(~month) +
  
  # Estética e Escalas
  theme_clean() +
  scale_fill_brewer(palette = "Set1") +
  expand_limits(y = max(df_letras_co2$y_pos) * 1.1) + # Garante espaço para a letra no topo
  labs(y = "log(fCO2 + 1)",
       x = "Cluster")


```

Testing CH4

```{r}
anova_ch4 <- aov(log(fch4+5) ~ cluster_pca*month, data=df_fluxTST)
summary(anova_ch4)

em_ch4 <- emmeans(anova_ch4, ~ cluster_pca | month)

pairs_ch4 <- pairs(em_ch4, adjust = "tukey")
summary(pairs_ch4)

cld_ch4 <- cld(em_ch4, Letters = letters, adjust = "tukey")
print(cld_ch4)
```

```{r}
df_letras_ch4 <- as.data.frame(cld_ch4) %>%
    mutate(.group = trimws(.group)) 


df_letras_ch4 <- df_letras_ch4 %>%
  mutate(
    erro_padrao = if("SE" %in% names(.)) SE else stderr,
    emmean = as.numeric(emmean),
    erro_padrao = as.numeric(erro_padrao)
  ) %>%
  group_by(month) %>%
  mutate(y_pos = max(emmean, na.rm = TRUE) + (2 * max(erro_padrao, na.rm = TRUE))) %>%
  ungroup()

```

```{r}
df_fluxTST %>% 
  mutate(month = factor(month, levels = c("set24", "nov24", "fev25", "jun25", "set25", "dez25"))) %>%
  ggplot(aes(x = cluster_pca, y = log(fch4 + 5))) +
  

  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(alpha = 0.3, width = 0.1, color = "grey40") +
  
  
  geom_text(data = df_letras_ch4, 
            aes(x = cluster_pca, y = y_pos, label = .group), 
            vjust = -0.5,           
            fontface = "bold", 
            size = 4, 
            color = "black") +
  

  facet_wrap(~month) +
  
  
  theme_clean() +
  scale_fill_brewer(palette = "Set1") +
  expand_limits(y = max(df_letras_ch4$y_pos) * 1.1) + 
  labs(y = "log(fCH4 + 5)",
       x = "Cluster")


ggsave("Plots/cluster_fluxtestch4.png", unit="cm", width = 20, height = 15)
```
