---
title: "Moisture Analysis"
author: "Larissa Verona"
date: "2025-01-14"
output: html_document
---

Moisture Data Processing - TOMST sensor

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pacotes
```{r}
rm(list=ls())
library(dplyr)
library(ggplot2)
library(myClim)
library(lubridate)
```

##Reading and processing data
```{r}

tms_file <- mc_read_files(
  "Data/Moisture",
  dataformat_name = "TOMST",
  recursive = FALSE,
  silent = TRUE
)
```

Checking the data
```{r}
mc_info_count(tms_file) #quantos sensores e medidas temos
mc_info_clean(tms_file) #limpando dados faltantes
mc_info(tms_file) %>% View()
mc_info_meta(tms_file) #sem metadados ainda
tms_file <- mc_prep_TMSoffsoil(tms_file) #out of soil measuraments
```

Adding metadata
```{r}
meta_data <- read.csv("Sensores_Metadata.csv", sep=";") %>% mutate(locality_id=as.character(locality_id))

str(meta_data)

tms_file <- mc_prep_meta_locality(tms_file, meta_data, param_name = NULL) #função para associar os metadados aos dados de fluxo. 
```

Selecting only last measuraments
```{r}
tms_file <- mc_join(tms_file, comp_sensors = NULL) # utilizo quando tenho mais de um arquivo relaciondo ao mesmo sensor
```

Cliping for interested time window
```{r}

#change according required time window
start <- as.POSIXct("2024-11-03", tz = "UTC")
end <- as.POSIXct("2025-12-31", tz = "UTC")

tms_file <- mc_prep_crop(tms_file, start, end)

#excluding measurements affected by fire on ING

# time window to exclude in ING
fire_start <- as.POSIXct("2025-03-06", tz = "UTC") # one day before fire
fire_end   <- as.POSIXct("2025-05-08", tz = "UTC") # one day after installing new sensors

# afected sensors
sensores_ing <- c("ING_p4a", "ING_p6a", "ING_p8a")
sensores_ing2 <- c("ING_p2c", "ING_p10b")
sensores_ing_ex <- c("ING_p2b", "ING_p4b",  "ING_p6b", "ING_p8b")


# non-afected sensors
localidades_todas <- names(tms_file$localities)
localidades_outros <- setdiff(localidades_todas, sensores_ing)
localidades_outros <- setdiff(localidades_outros, sensores_ing2)
localidades_outros <- setdiff(localidades_outros, sensores_ing_ex)

tms_outros <- tms_file %>% 
  mc_filter(localities = localidades_outros)

# ING data before fire
tms_ing_antes <- tms_file %>% 
  mc_filter(localities = sensores_ing) %>%
  mc_prep_crop(start, fire_start)

# ING data after fire
tms_ing_depois <- tms_file %>% 
  mc_filter(localities = sensores_ing) %>%
  mc_prep_crop(fire_end, end)

# ING P2c data (added after fire)
tms_ING_P2c <- tms_file %>% 
  mc_filter("ING_p2c") %>% mc_prep_crop(fire_end, end)

# ING P10b Data (added after fire, in the next field campaign. New sensor)

P10b_install  <- as.POSIXct("2025-06-04", tz = "UTC") #one day after P10b installing
tms_ING_P10b <- tms_file %>% 
  mc_filter("ING_p10b") %>% mc_prep_crop(P10b_install, end)

# Joing all
tms_file <- mc_prep_merge(list(tms_outros, tms_ing_antes, tms_ing_depois, tms_ING_P10b, tms_ING_P2c))
```
 
Correcting time
```{r}
tms_file_solar <- mc_prep_solar_tz(tms_file) #It doesn't seem to work

```

Converting moisture signal to volumetric soil water (according soil classes). Classes avaible at: https://ars.els-cdn.com/content/image/1-s2.0-S0168192318304118-mmc1.pdf

Previous approach involving categorization of soil classes in the field.
```{r}

#I'm not using this anymore. 
#tms_file <- mc_calc_vwc(tms_file, soiltype = "peat", localities = c("BBT_p10", "ING_p10a","ING_p10b", "7Q_p10", "VVT_p5", "VVT_p7", "VVT_p8", "CR_p9", "CR_p10"))

#tms_file <- mc_calc_vwc(tms_file, soiltype = "loamy sand A", localities = c("ING_p8a","ING_p8b", "ING_p4a","ING_p4b", "VVT_p1"))

#tms_file <- mc_calc_vwc(tms_file, soiltype = "sandy loam A", localities = c("ING_p2a", "ING_p2b", "ING_p2c", "ING_p6a", "ING_p6b","7Q_p1", "7Q_p3", "7Q_p5", "7Q_p7", "VVT_p4", "VVT_p10", "CR_p7"))

#tms_file <- mc_calc_vwc(tms_file, soiltype = "loam", localities = c("BBT_p7", "BBT_p8", "BBT_p5", "BBT_p3"))

#tms_file <- mc_calc_vwc(tms_file, soiltype = "sand", localities = c("CR_p3", "CR_p5"))
```

Current approach with classification based on soil data. See script MoistureSoilClasses.rmd
```{r}
class_soil <- read.csv("Data/Soils/Data_SoilClasses.csv")
class_soil <- class_soil[1:25,]

class_soil <- class_soil %>%
  mutate(ID = recode(ID,
                     "ING_p2" = "ING_p2a",
                     "ING_p4" = "ING_p4a",
                     "ING_p6" = "ING_p6a",
                     "ING_p8" = "ING_p8a",
                     "ING_p10" = "ING_p10a"))

class_soil <- bind_rows(
  class_soil,
  class_soil %>%
    filter(grepl("^ING_p(2|4|6|8|10)a$", ID)) %>%
    mutate(ID = sub("a$", "b", ID))
)

class_soil <- bind_rows(
  class_soil,
  class_soil %>%
    filter(ID == "ING_p2a") %>%
    mutate(ID = sub("a$", "c", ID)))

sand <- class_soil$ID[class_soil$classe == 1]
lsA <- class_soil$ID[class_soil$classe == 2]
lsB <- class_soil$ID[class_soil$classe == 3]
slA <- class_soil$ID[class_soil$classe == 4]
slB <- class_soil$ID[class_soil$classe == 5]
loam <- class_soil$ID[class_soil$classe == 6]
siltloam <- class_soil$ID[class_soil$classe == 7]
peat <- class_soil$ID[class_soil$classe == 8]

sand2 <- class_soil$ID[class_soil$classe2 == 1]
lsA2 <- class_soil$ID[class_soil$classe2 == 2]
lsB2 <- class_soil$ID[class_soil$classe2 == 3]
slA2 <- class_soil$ID[class_soil$classe2 == 4]
slB2 <- class_soil$ID[class_soil$classe2 == 5]
loam2 <- class_soil$ID[class_soil$classe2 == 6]
siltloam2 <- class_soil$ID[class_soil$classe2 == 7]
peat2 <- class_soil$ID[class_soil$classe2 == 8]

```

VWC calculation
```{r}
tms_file <- mc_calc_vwc(tms_file, soiltype = "sand", localities = sand)

tms_file <- mc_calc_vwc(tms_file, soiltype = "loamy sand A", localities = lsA)

tms_file <- mc_calc_vwc(tms_file, soiltype = "loamy sand B", localities = lsB)

tms_file <- mc_calc_vwc(tms_file, soiltype = "sandy loam A", localities = slA)

tms_file <- mc_calc_vwc(tms_file, soiltype = "sandy loam B", localities = slB)

tms_file <- mc_calc_vwc(tms_file, soiltype = "loam", localities = loam)

tms_file <- mc_calc_vwc(tms_file, soiltype = "silt loam", localities = siltloam)

tms_file <- mc_calc_vwc(tms_file, soiltype = "peat", localities = peat)
```


## Viewing

```{r}
tms_file %>% mc_filter(localities=c("VVT_p10", "VVT_p8", "VVT_p7", "VVT_p4", "VVT_p1")) %>% mc_plot_line(sensors = c("TMS_moist"))

tms_file %>% mc_filter(localities=c("VVT_p10", "VVT_p8", "VVT_p7", "VVT_p4", "VVT_p1")) %>% mc_plot_line(sensors = c("VWC_moisture"))

tms_file %>% mc_filter(localities=c("VVT_p10", "VVT_p8", "VVT_p7", "VVT_p4", "VVT_p1")) %>% mc_plot_raster(sensors = c("TMS_T1"))

tms_file %>% mc_filter(localities=c("CR_p10", "CR_p9", "CR_p7", "CR_p5", "CR_p3")) %>% mc_plot_line(sensors = c("TMS_moist"))

tms_file %>% mc_filter(localities=c("CR_p10", "CR_p9", "CR_p7", "CR_p5", "CR_p3")) %>% mc_plot_line(sensors = c("VWC_moisture"))

tms_file %>% mc_filter(localities=c("CR_p10", "CR_p9", "CR_p7", "CR_p5", "CR_p3")) %>% mc_plot_raster(sensors = c("TMS_T1"))

tms_file %>% mc_filter(localities=c("7Q_p10",  "7Q_p7", "7Q_p5", "7Q_p3", "7Q_p1")) %>% mc_plot_line(sensors = c("TMS_moist"))

tms_file %>% mc_filter(localities=c("7Q_p10",  "7Q_p7", "7Q_p5", "7Q_p3", "7Q_p1")) %>% mc_plot_line(sensors = c("VWC_moisture"))

tms_file %>% mc_filter(localities=c("7Q_p10",  "7Q_p7", "7Q_p5", "7Q_p3", "7Q_p1")) %>% mc_plot_raster(sensors = c("TMS_T1"))

tms_file %>% mc_filter(localities=c("ING_p10a", "ING_p10b", "ING_p8a",  "ING_p8b", "ING_p6a","ING_p6b", "ING_p4a", "ING_p4b", "ING_p2a", "ING_p2b", "ING_p2c")) %>% mc_plot_line(sensors = c("TMS_moist"))

tms_file %>% mc_filter(localities=c("ING_p10a", "ING_p10b", "ING_p8a",  "ING_p8b", "ING_p6a","ING_p6b", "ING_p4a", "ING_p4b", "ING_p2a", "ING_p2b", "ING_p2c")) %>% mc_plot_line(sensors = c("VWC_moisture"))

tms_file %>% mc_filter(localities=c("ING_p10",  "ING_p8", "ING_p6", "ING_p4", "ING_p2")) %>% mc_plot_raster(sensors = c("TMS_T1"))

tms_file %>% mc_filter(localities=c("BBT_p10",  "BBT_p8", "BBT_p7", "BBT_p5", "BBT_p3")) %>% mc_plot_line(sensors = c("TMS_moist"))

tms_file %>% mc_filter(localities=c("BBT_p10",  "BBT_p8", "BBT_p7", "BBT_p5", "BBT_p3")) %>% mc_plot_line(sensors = c("VWC_moisture"))

tms_file %>% mc_filter(localities=c("BBT_p10",  "BBT_p8", "BBT_p7", "BBT_p5", "BBT_p3")) %>% mc_plot_raster(sensors = c("TMS_T1"))
```


Daily aggregation

```{r}
tms_agg <- tms_file %>% mc_agg(tms_file, fun = c("mean"), period = "day", min_coverage = 0.8, use_utc=F)

tms_agg %>% mc_filter(localities=c("VVT_p10", "VVT_p8", "VVT_p7", "VVT_p4", "VVT_p1")) %>% mc_plot_line(sensors = c("VWC_moisture_mean"))

tms_agg %>% mc_filter(localities=c("CR_p10", "CR_p9", "CR_p7", "CR_p5", "CR_p3")) %>% mc_plot_line(sensors = c("VWC_moisture_mean"))

tms_agg %>% mc_filter(localities=c("7Q_p10",  "7Q_p7", "7Q_p5", "7Q_p3", "7Q_p1")) %>% mc_plot_line(sensors = c("VWC_moisture_mean"))

tms_agg %>% mc_filter(localities=c("ING_p10",  "ING_p8", "ING_p6", "ING_p4", "ING_p2")) %>% mc_plot_line(sensors = c("VWC_moisture_mean"))

tms_agg %>% mc_filter(localities=c("BBT_p10",  "BBT_p8", "BBT_p7", "BBT_p5", "BBT_p3")) %>% mc_plot_line(sensors = c("VWC_moisture_mean"))
```

## General analyses

Nice data, I can get interisting things.
```{r}
temp_env  <- mc_env_temp(tms_file, period = "all", min_coverage = 0.8)
moist_env <- mc_env_moist(tms_file, period = "all", min_coverage = 0.8)

```

## Getting a data frame
```{r}

tms_agg_hour <- tms_file %>% mc_agg(tms_file, fun = c("mean"), period = "hour", min_coverage = 0.25, use_utc=T)


tms_long<- mc_reshape_long(tms_file, use_utc = T) %>% mutate(sensor_name = as.factor(sensor_name), height = as.factor(height))  
tms_long <- tms_long %>% select(-c(serial_number, height))

write.csv(tms_long,"Data/Moisture/tms_data.csv")

levels(tms_long$height)
```
